---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Saima Arina"
date: "02/21/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r - loading csv}
staff <- read_csv("data/instructional-staff.csv")
```
```{r - pivoting}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```


### Exercise 1

```{r - line plot, fig.height=6, fig.width=11}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
   geom_line(size = 1.1) +
  labs(
    title = "Trends in Instructional Staff Employment, 1975–2011",
    x = "Year",
    y = "Percent of instructional staff",
    color = "Faculty type"
  ) +
  theme_minimal(base_size = 15)
```



### Exercise 2

To highlight an increase in the proportion of part-time faculty over time compared to other staff types, I would put the visual emphasis on part-time faculty through a bolder color and muting the colors of the other types. I could also increase the line thickness for part-time and decrease it for the others. 

### Exercise 3

```{r - loading}
fisheries <- read_csv("data/fisheries.csv")
```

To start with, the 3D pie charts are not visually appealing and do not provide comparisons equally across countries. Bar charts can replace them to make the areas and lenghts comparable. Putting capture and aquaculture side by side within each country and differentated by one color per production type instead of per country can show comparison and the total. The rank pattern can be made more obvious by ordering the countries by either total production or capture. The visuals are also missing a title, axis labels, and an additional subtitle. 



```{r - pivot}
fisheries_long <- fisheries %>%
  pivot_longer(cols = c(capture, aquaculture), names_to = "type", values_to = "tons")
```

```{r - plotting}
top_countries <- fisheries_long %>% 
  group_by(country) %>% 
  summarise(total_tons = sum(tons, na.rm = TRUE)) %>% 
  slice_max(total_tons, n = 15) %>% 
  arrange(desc(total_tons)) %>% 
  pull(country)

fisheries_top15 <- fisheries_long %>% 
  filter(country %in% top_countries) %>% 
  mutate(country = factor(country, levels = top_countries))

ggplot(fisheries_top15,
       aes(x = country,
           y = tons,
           fill = type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Capture vs Aquaculture, Top 15 Producing Countries (2016)",
    x = "Country",
    y = "Tons of fish (2016)"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Smokers in Whickham 

```{r}
data(Whickham)
?Whickham
library(performance)
```

### Exercise 1

This study is observational because there was no experimental manipulation done on the participants. Only observations were recorded (whether they smoke and their age). 


### Exercise 2

```{r - obs}
nrow(Whickham)
```
There are 1,314 observations and each observation is one individual person from Whickham with their recorded age, smoking status at baseline, and survival status 20 years later.


### Exercise 3
```{r - variables}
ncol(Whickham)
```

There are 3 variables: outcome (categorical), smoker (categorical), and age (numerical). 


```{r - variable visuals}
Whickham %>% 
  ggplot(aes(x = smoker)) +
  geom_bar()

Whickham %>% 
  ggplot(aes(x = outcome)) +
  geom_bar()

Whickham %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5)
```

### Exercise 4

I would imagine there to be an association between being a smoker and having a higher likelihood of death.


### Exercise 5

```{r - count}
Whickham %>%
  count(smoker, outcome)
```
```{r - visual bar chart}
Whickham %>%
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Health outcome by smoking status in Whickham study",
    x = "Smoking status",
    y = "Proportion within smoking group",
    fill = "Outcome"
  ) +
  theme_minimal()
```

Conditional Probabilities

Non-smokers: 502 alive, 230 dead = 732
P(Dead|Non-smoker) = 230/732 = 0.31 or 31%
P(Alive|Non-smoker) = 502/732 = 0.69 or 69%

Smokers: 443 alive, 139 dead = 582
P(Dead|Smoker) = 139/582 = 0.24 or 24%
P(Alive|Smoker) = 443/582 = 0.76 or 76%

This is conflicting with my initial assumption, since it's found that smokers actually have a lower observed death proportion and a higher survival proportion than non‑smokers.


### Exercise 6

```{r - binning age}
Whickham <- Whickham %>%
  mutate(
    age_cat = case_when(
      age <= 44 ~ "18-44",
      age > 44 & age <= 64 ~ "45-64",
      age > 64 ~ "65+"
    )
  )
```


### Exercise 7 

```{r - new plot}
Whickham %>%
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ age_cat) +
  labs(
    title = "Health outcome by smoking status, faceted by age group",
    x = "Smoking status",
    y = "Proportion within smoking group",
    fill = "Outcome"
  ) +
  theme_minimal()
```

```{r - contingency}
Whickham %>%
  count(smoker, age_cat, outcome)
```


What changed was that now smokers are found to have a higher proportion of death than non-smokers, and this was prominent in the older age categories, although overall smokers seemed to be healthier in the plot that was unfaceted. In the extended table, we see that younger people have more lower death rates overall and there is an age imbalance. It is after stratifying by ages that we can see the harmful association we expect between death and smoking.




