---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Saima Arina"
date: "02/21/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r - loading csv}
staff <- read_csv("data/instructional-staff.csv")
```
```{r - pivoting}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```


### Exercise 1

```{r - line plot, fig.height=6, fig.width=11}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
   geom_line(size = 1.1) +
  labs(
    title = "Trends in Instructional Staff Employment, 1975–2011",
    x = "Year",
    y = "Percent of instructional staff",
    color = "Faculty type"
  ) +
  theme_minimal(base_size = 15)
```



### Exercise 2

To highlight an increase in the proportion of part-time faculty over time compared to other staff types, I would put the visual emphasis on part-time faculty through a bolder color and muting the colors of the other types. I could also increase the line thickness for part-time and decrease it for the others. 

### Exercise 3

...

Add exercise headings as needed.
