---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Saima Arina"
date: "02/21/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r - loading csv}
staff <- read_csv("data/instructional-staff.csv")
```
```{r - pivoting}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```


### Exercise 1

```{r - line plot, fig.height=6, fig.width=11}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
   geom_line(size = 1.1) +
  labs(
    title = "Trends in Instructional Staff Employment, 1975–2011",
    x = "Year",
    y = "Percent of instructional staff",
    color = "Faculty type"
  ) +
  theme_minimal(base_size = 15)
```



### Exercise 2

To highlight an increase in the proportion of part-time faculty over time compared to other staff types, I would put the visual emphasis on part-time faculty through a bolder color and muting the colors of the other types. I could also increase the line thickness for part-time and decrease it for the others. 

### Exercise 3

```{r - loading}
fisheries <- read_csv("data/fisheries.csv")
```

To start with, the 3D pie charts are not visually appealing and do not provide comparisons equally across countries. Bar charts can replace them to make the areas and lenghts comparable. Putting capture and aquaculture side by side within each country and differentated by one color per production type instead of per country can show comparison and the total. The rank pattern can be made more obvious by ordering the countries by either total production or capture. The visuals are also missing a title, axis labels, and an additional subtitle. 



```{r - pivot}
fisheries_long <- fisheries %>%
  pivot_longer(cols = c(capture, aquaculture), names_to = "type", values_to = "tons")
```

```{r - plotting}
top_countries <- fisheries_long %>% 
  group_by(country) %>% 
  summarise(total_tons = sum(tons, na.rm = TRUE)) %>% 
  slice_max(total_tons, n = 15) %>% 
  arrange(desc(total_tons)) %>% 
  pull(country)

fisheries_top15 <- fisheries_long %>% 
  filter(country %in% top_countries) %>% 
  mutate(country = factor(country, levels = top_countries))

ggplot(fisheries_top15,
       aes(x = country,
           y = tons,
           fill = type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Capture vs Aquaculture, Top 15 Producing Countries (2016)",
    x = "Country",
    y = "Tons of fish (2016)"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Smokers in Whickham 

```{r}
data(Whickham)
?Whickham
library(performance)
```

### Exercise 1

This study is observational because there was no experimental manipulation done on the participants. Only observations were recorded (whether they smoke and their age). 


### Exercise 2

```{r - obs}
nrow(Whickham)
```
There are 1,314 observations and each observation is one individual person from Whickham with their recorded age, smoking status at baseline, and survival status 20 years later.


### Exercise 3
```{r - variables}
ncol(Whickham)
```

There are 3 variables: outcome (categorical), smoker (categorical), and age (numerical). 


```{r - variable visuals}
Whickham %>% 
  ggplot(aes(x = smoker)) +
  geom_bar()

Whickham %>% 
  ggplot(aes(x = outcome)) +
  geom_bar()

Whickham %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5)
```

### Exercise 4

I would imagine there to be an association between being a smoker and having a higher likelihood of death.


